---
title: "Covid-viz"
author: "Josephat Oyondi"
date: "2023-01-22"
output: html_document
runtime: shiny
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Loading libraries
```{r warning=FALSE}
library(plotly)
library(dplyr)
library(magrittr)
library(ggplot2)
library(countrycode)
library(lubridate)
Sys.setenv(MAPBOX_TOKEN = 11122223333444) #presevents the mapbox token error
```

```{r}
data = read.csv('owid-covid-data.csv')

#head(data)
```
## Data preprocessing, checking on the column names and determine the important ones to help me in achieving the objectives, checking for null/NAN values.


## Write the column names to a text file for my reference in the analysis
```{r}
columns <- colnames(data)
file_columns<-file("columns.txt")
writeLines(c(columns), file_columns)
close(file_columns) #close the file
```

## Check columns with the null values
```{r}
#which(is.na(data))
#cols_with_na <- which(apply(data, 2, function(x) any(is.na(x))))
#colnames(data)[cols_with_na] #this indicates that atleast there are missing records in one or more rows in each feature included in the dataset.Except iso_code, continent, location and date
#colSums(is.na(data))
```
## Visualizing the total reported cases since the start of the pandemic
```{r warning=FALSE}
df <- data
# show difference between paths and lines
p <- df %>%
  arrange(total_cases) %>%
  plot_ly(x = ~date, y = ~total_cases)
add_lines(p)


```
### Considering the new tests
```{r}
p <- df %>%
  arrange(new_tests) %>%
  plot_ly(x = ~date, y = ~new_tests) %>% 
add_lines(p)
```

## A notable issue for the visualization is that although there were millions of reported cases, the number of new tests was relatively low, implying that not all countries could have accounted to new tests but had highest number of reported cases. 
# To confirm the validity of this conclusion, the following visualization digs deep into highlighting the top countries with most cases and most new tests

```{r warning=FALSE}
#group the dataset into years and months
dates <- c(df$date)
months <- month(ymd(dates))
years <- year(ymd(dates))
#new df
new_df <- data.frame(Month = months, Year = years, Continent = df$continent, total_cases = df$total_cases)

#create a plot for top 10 continents
top10 <- new_df %>% 
  group_by(Continent, Year) %>%
  summarise(totals = sum(total_cases, na.rm = TRUE)) %>%
  filter(!is.na(Continent)) %>%
  arrange(desc(totals)) %>%
  top_n(10) 

top10 %>%
  plot_ly(x = ~Continent, y = ~totals, type = "bar")

```

```{r}
#use the new datasets with fixed latitudes for better map visualization.
new_data <- read.csv("coronavirus.csv")
save(new_data, file = "corona.RData")

```

```{r warning=FALSE}
#get the leading countries in terms of the number of cases confirmed

filtered_df <- new_data %>% 
  filter(type == "confirmed") %>%
  group_by(country) %>%
  summarise(sum_Cases = sum(cases)) %>%
  arrange(-sum_Cases)%>%
  top_n(10) #order the total in ascending order

plt <- ggplot(filtered_df, aes(x = country, y=sum_Cases)) + geom_bar(stat="identity", fill="steelblue")+
  theme_minimal()
ggplotly(plt)


```

```{r}
load("corona.RData")
corona_data <- new_data

head(corona_data)

```



```{r warning=FALSE}
library(shiny)

#creating a chroloplot
dates = c(corona_data$date)
Date = ymd(dates)

#new_data$code3 <- countrycode(new_data$country, "country.name", "iso3c") #convert the names into 3 letter country codes

cases_summed <- corona_data %>% 
  mutate(Year = year(Date), Month = month(Date)) %>%
  mutate(YearMonth = paste(format(as.Date(paste(Month, 1, Year), "%m %d %Y"), "%b %Y"), sep = " ")) %>%
  group_by(country, YearMonth, iso3) %>% 
  filter(cases > 0) %>% #used to filter all cases recorded in negative
  summarise(cases_sum = sum(cases))


#cases seem to have been recorded with negative values

#get minimum and maximum values
min_value <- min(cases_summed$cases_sum)
max_value <- max(cases_summed$cases_sum)

cat("Minimum value in cases_sum column:", min_value, "\n")
cat("Maximum value in cases_sum column:", max_value, "\n")

# Define a function to assign colors to countries
assign_colors <- function(country) {
  colors <- c("red", "blue", "green", "yellow", "purple")
  country_colors <- ifelse(country %in% c("country1", "country2", "country3"), colors[1], 
                          ifelse(country %in% c("country4", "country5", "country6"), colors[2],
                          ifelse(country %in% c("country7", "country8", "country9"), colors[3],
                          ifelse(country %in% c("country10", "country11", "country12"), colors[4],
                          ifelse(country %in% c("country13", "country14", "country15"), colors[5], 
                          "grey")))))
  return(country_colors)
}

# Define the UI for the Shiny app
ui <- fluidPage(
  titlePanel("Cases by Country"),
  sidebarLayout(
    sidebarPanel(
      selectInput("month", "Month:", c("All", unique(cases_summed$YearMonth)))
     
    ),
    mainPanel(
      plotlyOutput("choropleth")
    )
  )
)

# Define the server logic for the Shiny app
server <- function(input, output) {
  
  # Filter the data based on the selected month and year
  filtered_data <- reactive({
    if (input$month == "All") {
      cases_summed
    } else {
      cases_summed %>% filter(YearMonth == input$month)
    }
  })
  
  # Plot the choropleth map
  output$choropleth <- renderPlotly({
    filtered_data() %>%
      mutate(color = assign_colors(country)) %>%
      plot_ly(z = ~cases_sum, text = ~country, locations = ~iso3, locationmode = "ISO-3", type = "choropleth",
              color = ~color) %>%
      colorbar(title = "Total Covid-reported cases ") %>%
      layout(title = "Cases by Country", geo = list(showframe = FALSE, showcoastlines = FALSE))
  })
  
}

# Run the Shiny app
shinyApp(ui = ui, server = server,  options = list(width = "100%", height = "400px"))
```

### Creating a new choropleth for the cases using a slider input
 
```{r warning=FALSE}
library(shiny)

# Read the data file into a data frame
corona_data <- read.csv("coronavirus.csv")
#confirmed
confirmed <- corona_data %>% 
  filter(type == "confirmed") %>%
  group_by(date, country, iso3) %>%
  summarise(total_cases = sum(cases)) %>%
  arrange(-total_cases)
#deaths
deaths <- corona_data %>% 
  filter(type == "death") %>%
  group_by(date, country, iso3) %>%
  summarise(total_cases = sum(cases)) %>%
  arrange(-total_cases)
#recovered
recovered <- corona_data %>% 
  filter(type == "recovery") %>%
  group_by(date, country, iso3) %>%
  summarise(total_cases = sum(cases)) %>%
  arrange(-total_cases)
# Define the user interface of the Shiny app
ui <- fluidPage(
  # Create a header with a title panel
  navbarPage("COVID-19 Dashboard",
    tabPanel("Map",
      sidebarLayout(
        sidebarPanel(
          selectInput(inputId = "case_type", label = "Choose type of cases to analyze:",
                      choices = c("All", "Confirmed", "Died", "Recovered")),
          sliderInput(
            "slider", "Time",
            # Set the minimum value of the slider to the earliest date in the data
            min = as.Date(corona_data$date[1]),
            # Set the maximum value of the slider to the latest date in the data
            max = as.Date(corona_data$date[length(corona_data$date)]),
            # Set the default value of the slider to the latest date in the data
            value = as.Date(corona_data$date[length(corona_data$date)]),
            timeFormat = "%b %Y"
          )
        ),
        mainPanel(
          plotlyOutput("choropleth")
        )
      )
    ),
    tabPanel("Charts",
      # Add charts here
    ),
    tabPanel("Table",
      # Add table here
    )
  )
)


# Define the server of the Shiny app
server <- shinyServer(function(input, output, session){
  # Define a reactive expression to subset the data and calculate the sum of cases by country for the selected date
  #define different datasets for confirmed, deaths and recovered
  

  
  selected_data <- reactive({
    selected_date <- as.Date(input$slider)
    
      #logic for specifying the kind of data to be used in the choropleth
    # Check which value was selected using an if statement
    if (input$case_type == "All") {
      # Do something if "All" was selected
      subset_data <- corona_data[corona_data$date == selected_date, ]
      sum_cases <- aggregate(subset_data$cases, 
                           by = list(country = subset_data$country, iso3 = subset_data$iso3), 
                           FUN = sum)
    } else if (input$case_type == "Confirmed") {
      # Do something if "Confirmed" was selected
      subset_data <- confirmed[confirmed$date == selected_date, ]
      sum_cases <- aggregate(subset_data$total_cases, 
                           by = list(country = subset_data$country, iso3 = subset_data$iso3), 
                           FUN = sum)
    } else if (input$case_type == "Died") {
      subset_data <- deaths[deaths$date == selected_date, ]
      sum_cases <- aggregate(subset_data$total_cases, 
                           by = list(country = subset_data$country, iso3 = subset_data$iso3), 
                           FUN = sum)
    } else if (input$case_type == "Recovered") {
      subset_data <- recovered[recovered$date == selected_date, ]
      sum_cases <- aggregate(subset_data$total_cases, 
                           by = list(country = subset_data$country, iso3 = subset_data$iso3), 
                           FUN = sum)
    }
    
    
    return(sum_cases)
  })

  
  # Define the output of the choropleth map as a plotly object that shows a choropleth map of the cases by country
  output$choropleth <- renderPlotly({
    selected_data() %>%
      plot_ly(z = selected_data()$x, text = ~selected_data()$country, 
              locations = ~selected_data()$iso3, 
              locationmode = "ISO-3", type = "choropleth", 
              color = "Bergeron") %>%
      colorbar(title = "Total Covid-reported cases ") %>%
      layout(title = "Cases by Country", 
             geo = list(showframe = FALSE, showcoastlines = FALSE))
  })
})

# Run the Shiny app
shinyApp(ui = ui, server = server)


```
```{r}

```
